Cartogram Mapping and its Application to Cancer Data Visualisation

### 2019-09-10 review_article.Rmd
Section links

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, fig.width = 6, fig.height = 5)
library(ggplot2)
library(grid)
library(png)
library(gridExtra)
```
<!--

@@

<!-- Structure of the paper -->
The paper is organised as follows.  The next section describes the choropleth map which is the common approach to disease maps. Section XXX surveys atlases in use today. Section XXX describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section XXX. Disease maps are ore useful when made interactive, and common options are described in Section XXX, along with a discussion of benefits and disadvantages. The last section summarises the paper and discusses future directions.

# Map displays for disease data

@@

```{r choropleth-grid-create, message=FALSE, fig.cap = "Public facing choropleth cancer maps.[a. The Environment and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/) (Lung cancer in females)  (Incidence rates per 100,000 for all cancers, male and female), [b. Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx),  [c. Atlas of Cancer in QLD - Relative incidence ration, Lung cancer in males](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/), [d. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), [e. United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/) (Incidence rate per 100,000, all cancer types, men and women), [f. Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) (Relative risk, Lung cancer, men vs women), [g. An Atlas of Cancer in South Australia](https://www.cancersa.org.au/assets/images/pdfs/An%20Atlas%20of%20Cancer%20in%20South%20Australia%20-%20Full%20Report.pdf) (Incidence rate per 100,000, lung cancer in females) "}
library(cowplot)

@@

ggdraw() +
  draw_plot(rasterGrob(plot1), 0, .75, 0.33, .25) +
  draw_plot(rasterGrob(plot2), 0.33, 0.75, 0.33, .25) +
  draw_plot(rasterGrob(plot3), 0.66, 0.5, .33, .5) +
  draw_plot(rasterGrob(plot4), 0, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot5), 0.33, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot6), 0, 0, .66, .5) +
  draw_plot(rasterGrob(plot7), .66, 0, .33, .5) +
  draw_plot_label(c("a", "b", "c", "d", "e", "f", "g"), 
    c(0.015, 0.33, 0.63, 0.015, 0.33, 0.014, 0.63), 
    c(0.99, 0.99, 0.99, 0.74, 0.74, 0.49, 0.49), size = 15)

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses. These visualisation enable thse non-expert audiences to access the outputs of sophisticated statistical analyses. Cruickshank's (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly cancer maps utilised to communicate to the public and other non-expert audiences are choropleths. A grey literature review of cancer atlases available in the public domain [@roberts2019communication] identified 33 cancer atlases published on the internet between January 2010 and November 2015, all of which were choropleths. These maps were published by non-commercial organisations, including not-for-profits, government, research organisations, advocacy groups or government funded partnerships.Only one map was published by a commercial entity.  The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases covered geographies from all around the world. One atlas covered the European Union. Most focused on single nations, the United States was considered by eleven atlases, the United Kingdom by seven, followed by three of Australia, two of Canada, and one of each from Switzerland, Germany, Norway. 
Figure 2.a.b and c show a global, national and state choropleth respectively. 
Not all maps had a national focus and ten covered a region or state rather than an entire nation. The states or counties/regions covered were South Australia (AUS), Queensland (AUS), Ontario (CAN), Valencia (Spain), Pennsylvania county Massachusetts (US), New Hampshire (US), Cape Cod (US), Missouri (US), Florida (US), New York State (US) and Arizona (US).

@@

# Alternative displays

@@

```{r choroCRS, fig.cap = "Four choropleth maps of the United States of America using various coordinate reference systems. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The map projections alter the shapes and angles of the boundaries of each state. Maps a and b are similar in their straight edges, unlike maps c and d which curve on the northern United States border.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggchoroCRS.png")
```

@@

```{r ggdorl, fig.cap = "A dorling cartogram of the Unites States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbours, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.", message=FALSE, warning=FALSE, out.width='80%'}
knitr::include_graphics("figures/ggdorl.png")
```

@@

# Comparison of mapping methods

@@

# Animation and Interactivity

>>>>

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, cache = TRUE, fig.height = 5)
library(ggplot2)
library(grid)
library(png)
library(gridExtra)
```
<!--

@@

<!-- Structure of the paper -->
The paper is organised as follows.  The next section, [Map displays for disease data](#choropleths), describes the choropleth map which is the common approach to disease maps. Section [Publicly available atlases](#public) surveys atlases in use today. The [Alternative displays](#alternative) describes an alternative display, the cartogram what may be useful for countries that have heterogeneously sized geographic units. The pros and cons of these approaches is discussed in Section [Comparison of mapping methods](#compare). Disease maps are more useful when made interactive, and common options are described in Section [Animation and Interactivity](#interacting), along with a discussion of benefits and disadvantages. The last section summarises the paper and discusses future directions.

# Map displays for disease data {#choropleths}

@@

```{r choropleth-grid-create, message=FALSE, fig.cap = "Public facing choropleth cancer maps.[a. The Environment and Health Atlas of England and Wales](http://www.envhealthatlas.co.uk/eha/Breast/) (Lung cancer in females)  (Incidence rates per 100,000 for all cancers, male and female), [b. Globocan 2012: Estimated  Cancer Incidence, Mortality and Prevalence](http://globocan.iarc.fr/Pages/Map.aspx),  [c. Atlas of Cancer in QLD - Relative incidence ration, Lung cancer in males](https://cancerqld.org.au/research/queensland-cancer-statistics/queensland-cancer-atlas/), [d. Bowel Cancer Australia Atlas](http://www.bowelcanceratlas.org/), [e. United States Cancer Statistics: An Interactive Cancer Statistics Website](https://nccd.cdc.gov/DCPC_INCA/) (Incidence rate per 100,000, all cancer types, men and women), [f. Map of Cancer Mortality Rates in Spain](http://elpais.com/elpais/2014/10/06/media/1412612722_141933.html) (Relative risk, Lung cancer, men vs women), [g. An Atlas of Cancer in South Australia](https://www.cancersa.org.au/assets/images/pdfs/An%20Atlas%20of%20Cancer%20in%20South%20Australia%20-%20Full%20Report.pdf) (Incidence rate per 100,000, lung cancer in females)"}
library(cowplot)

@@

ggdraw() +
  draw_plot(rasterGrob(plot1), 0, .75, 0.33, .25) +
  draw_plot(rasterGrob(plot2), 0.33, 0.75, 0.33, .25) +
  draw_plot(rasterGrob(plot3), 0.66, 0.5, .33, .5) +
  draw_plot(rasterGrob(plot4), 0, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot5), 0.33, 0.5, .33, .25) +
  draw_plot(rasterGrob(plot6), 0, 0, .66, .5) +
  draw_plot(rasterGrob(plot7), .66, 0, .33, .5) +
  draw_plot_label(c("a", "b", "c", "d", "e", "f", "g"), 
    c(0.015, 0.33, 0.63, 0.015, 0.33, 0.014, 0.63), 
    c(0.98, 0.98, 0.98, 0.74, 0.74, 0.47, 0.47), size = 15)

@@

Cancer maps are effective tools for communicating to wide range of audiences, including the general public and others not trained in statistical analyses. These visualisation enable thse non-expert audiences to access the outputs of sophisticated statistical analyses. Cruickshank's (1947) as cited by @DMAHP, discusses using visuals as a 'formal statistical assessment of the spatial pattern'. Overwhelmingly cancer maps utilised to communicate to the public and other non-expert audiences are choropleths. A grey literature review of cancer atlases available in the public domain [@roberts2019communication] identified 33 cancer atlases published on the internet between January 2010 and November 2015, all of which were choropleths. These maps were published by non-commercial organisations, including not-for-profits, government, research organisations, advocacy groups or government funded partnerships.Only one map was published by a commercial entity.  The use of choropleths within the public domain mirrors the heavy use of choropleth maps within the research literature, discussed above. 
The cancer atlases covered geographies from all around the world, most focused on single nations.
Figure \@ref(choropleth-grid-create) displays a global (b), national (a,d,e,f) and state (c,g) choropleths.

@@

# Alternative displays {#alternative}

@@

```{r choroCRS, fig.cap = "Four choropleth maps of the United States of America using various coordinate reference systems. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. The map projections alter the shapes and angles of the boundaries of each state. Maps a and b are similar in their straight edges, unlike maps c and d which curve on the northern United States border.", message=FALSE, warning=FALSE, out.width='80%', fig.align='center', fig.width=12}
knitr::include_graphics("figures/ggchoroCRS.png")
```

@@

```{r ggdorl, fig.cap = "A dorling cartogram of the Unites States of America. Each state has been coloured according to the average age-adjusted rate of incidence for lung and bronchus for females and males in the United States 2012-2016. Each state has been represented by a circle, but the size was determined by the population of the state in 2015. The North East states remain closer to their neighbours, and may be displaced from their geographic location. The sparsity of the population in the Mid West is highlight by the distance between the circles, located at the geographic centroids.", message=FALSE, warning=FALSE, out.width='80%', fig.align='center', fig.width=12}
knitr::include_graphics("figures/ggdorl.png")
```

@@

# Comparison of mapping methods (#compare)

@@

# Animation and Interactivity {#interacting}
